---
title: "Take Home Exercise 2"
date: "6 December 2023"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  warning: false
editor: visual
---

## Getting Started

```{r}
pacman::p_load(tmap,sf,spdep,sfdep,tidyverse,knitr,plotly,zoo,Kendall)
```

## Preparing the Flow Data

### Importing the OD data

Firstly, we will import the **Passenger Volume by Origin Destination Bus Stops** data set downloaded from LTA DataMall by using read_csv() of **readr** package.

```{r}
odbus <- read_csv("data/aspatial/origin_destination_bus_202308.csv")
```

Check odbus tibble data frame that values in OROGIN_PT_CODE and DESTINATION_PT_CODE are in numeric data type.

```{r}
glimpse(odbus)
```

Origin & Destination Bus Stop Code

```{r}
odbus$ORIGIN_PT_CODE <-
as.factor(odbus$ORIGIN_PT_CODE)
odbus$DESTINATION_PT_CODE <-
as.factor(odbus$DESTINATION_PT_CODE)
```

### Extracting the Study Data

Filter out data that belong to trips that occur on:

-   "Weekday" and "6-9am" (wdmp)

    ```{r}
    wdmp <-  odbus %>%
      filter(DAY_TYPE == "WEEKDAY") %>%
      filter(TIME_PER_HOUR >= 6 &
               TIME_PER_HOUR <= 9) %>%
      group_by(ORIGIN_PT_CODE) %>%
      summarise(TRIPS = sum(TOTAL_TRIPS))
    ```

Check resulting data tables

```{r}
kable(head(wdmp)) 
```

Output saved in rds format for future use

```{r}
write_rds(wdmp, "data/rds/wdmp.rds") 
```

Import the rds file into R environment

```{r}
wdmp <- read_rds("data/rds/wdmp.rds") 
```

## Working with Geospatial Data

Two geospatial data (shapefile) will be used for this exercise:

-   BusStop: Provides location of bus stop as at Q4 2022

-   MPSZ-2019: This data provides the sub-zone boundary of URA Master Plan 2019

### Importing geospatial data

```{r}
busstop <- st_read(dsn = "Data/geospatial",
                   layer = "BusStop") %>%
  st_transform(crs = 3414)

mpsz <- st_read(dsn = "data/geospatial",
                layer = "MPSZ-2019") %>%
  st_transform(crs = 3414)
```

Check structure of `busstop` and `MPSZ`  sf tibble data frame

```{r}
glimpse(busstop)
glimpse(mpsz)
```

## Geospatial Data Wrangling

### Combining Busstop & mpsz

```{r}
busstop_mpsz <- st_intersection(busstop, mpsz) %>%
  select(BUS_STOP_N, SUBZONE_C) %>%
  st_drop_geometry()
```

Save output into rds format

```{r}
write_rds(busstop_mpsz, "data/rds/busstop_mpsz.csv")
```

## Hexagon Grid Setup

### Drawing the Hexagon Grid

Draw hexagon grid over the mpsz map

```{r}
area_honeycomb_grid = st_make_grid(mpsz, c(750, 750), what = "polygons", square = FALSE)
```

Convert the hexagon grid to sf (simple features) object and add a new column grid_id (sequential identifier) to it

```{r}
honeycomb_grid_sf = st_sf(area_honeycomb_grid) %>%
  # add grid ID
  mutate(grid_id = 1:length(lengths(area_honeycomb_grid)))
```

Determine which bus stops is contained within which hexagon using **`st_within`**.

```{r}
busstop_honeycomb <- st_intersection(honeycomb_grid_sf,busstop) %>%
  select(BUS_STOP_N, grid_id) %>%
  st_drop_geometry()
```

Save output into rds format

```{r}
write_rds(busstop_honeycomb, "data/rds/busstop_honeycomb.csv")
```

Check for duplicate records

```{r}
duplicate <- busstop_honeycomb %>%
  group_by_all() %>%
  filter(n()>1) %>%
  ungroup()
```

Retain unique records only

```{r}
busstop_honeycomb <- unique(busstop_honeycomb)
```

Only include hexagon grid IDs that have bus stop numbers

```{r}

busstop_honeycomb <- busstop_honeycomb %>%
  filter(!is.na(grid_id) & grid_id > 0)
```

### Assign Each Bus Stop to a Grid ID

For all time periods, assign bus stops to a hexagon grid id. All NULL and 0 values are removed.

```{r}
wdmp_gridid <- left_join(busstop_honeycomb, wdmp,
            by = c("BUS_STOP_N" = "ORIGIN_PT_CODE")) 

wdmp_gridid <- wdmp_gridid %>%
  filter(!is.na(TRIPS) & TRIPS > 0)


wdap_gridid <- left_join(busstop_honeycomb, wdap,
            by = c("BUS_STOP_N" = "ORIGIN_PT_CODE")) 

wdap_gridid <- wdap_gridid %>%
  filter(!is.na(TRIPS) & TRIPS > 0)


hmp_gridid <- left_join(busstop_honeycomb, hmp,
            by = c("BUS_STOP_N" = "ORIGIN_PT_CODE")) 

hmp_gridid <- hmp_gridid %>%
  filter(!is.na(TRIPS) & TRIPS > 0)


hep_gridid <- left_join(busstop_honeycomb, hep,
            by = c("BUS_STOP_N" = "ORIGIN_PT_CODE")) 

hep_gridid <- hep_gridid %>%
  filter(!is.na(TRIPS) & TRIPS > 0)

```

Note: Grid id 2955 and 3036 contains bus stops that are not within Singapore (i.e. Malaysia). These are left as they could provide insights to traffic between the Singapore & Malaysia land border.

## **Choropleth Visualisation**

#### **Weekday Morning Peak 6am-9am**

Sum up the trips per hexagon

```{r}
total_trips_by_grid_wdmp <- wdmp_gridid %>%
  group_by(grid_id) %>%
  summarise(total_trips = sum(TRIPS, na.rm = TRUE))

```

Merge geospatial data

```{r}

total_trips_by_grid_wdmp <- total_trips_by_grid_wdmp %>%
  left_join(honeycomb_grid_sf, by = c("grid_id" = "grid_id"))

total_trips_by_grid_wdmp_sf <- st_sf(total_trips_by_grid_wdmp)
```

Summary Statistics (Hexagon)

```{r}
summary(wdmp$TRIPS)
```

Total Trips in the time slot

```{r}
sum(wdmp$TRIPS)
```

Plot the Choropleth map

```{r}

tmap_mode("plot")

tm_shape(total_trips_by_grid_wdmp_sf) +
  tm_fill(
    col = "total_trips",
    palette = "Greens",
    style = "cont",
    title = "Total Trips Taken - Weekday Morning Peak 6-9am",
    id = "grid_id",
    showNA = FALSE,
    alpha = 0.6,
    popup.vars = c(
      "Number of trips: " = "total_trips"
    ),
    popup.format = list(
      total_trips = list(format = "f", digits = 0)
    )
  ) +
  tm_borders(col = "grey40", lwd = 0.4)
```

::: {.callout-tip title="Description of Spatial Patterns" style="color: green"}
Weekday Morning Peak 6am-9am

This timeslot has a total of 26,430,413 trips taken. It ranks first by total volume of trips among our time periods of interest. The median number of trips per hexagon is 2184.

Hexagons with greater number of trips are located near:

-   Residential areas e.g. Tampines, Ang Mo Kio

-   Public transport hubs e.g. Bus Interchange, MRT Interchange e.g. Toa Payoh Interchange, Bedok Interchange

-   Immigration checkpoint at Woodlands
:::



## **Global Spatial Autocorrelation**

Computed to perform spatial complete randomness test for global spatial autocorrelation.

### **Computing Contiguity Spatial Weights**

Use poly2nb() of spdep package to compute contiguity weight matrices for the study area. This function builds a neighbours list based on regions with contiguous boundaries.

#### **Weekday Morning Peak 6am-9am**

```{r}
wm_wdmp <- poly2nb(total_trips_by_grid_wdmp_sf, 
                queen=TRUE)
summary(wm_wdmp)
```

Note: The summary report above shows that there are 1490 hexagons considered for Weekday Morning Peak 6am-9am. The most connected hexagon has 6 neighbours. There are 43 hexagons with only 1 neighbour.



### Computing Monte Carlo Moran's I Statistic

Seed 1234 is used for consistency and a total of 1000 simulations will be performed

#### **Weekday Morning Peak 6am-9am**

```{r}
set.seed(1234)

bperm_wdmp = moran.mc(total_trips_by_grid_wdmp_sf$total_trips,
                      listw = rswm_wdmp,
                      nsim = 999,
                      zero.policy = TRUE,
                      na.action = na.omit)
bperm_wdmp
```


## **Cluster and Outlier Analysis**

Evaluare existence of clusters in the spatial arrangement of trips in time periods of interest e.g. Weekday Morning Peak 6am-9am.

### Computing Local Moran's I

#### **Weekday Morning Peak 6am-9am**

```{r}
localMI_wdmp <- localmoran(total_trips_by_grid_wdmp_sf$total_trips, rswm_wdmp)
head(localMI_wdmp)
```



Note: localmoran() function returns a matrix of values whose columns are:

-   Ii: the local Moran's I statistics

-   E.Ii: the expectation of local Moran statistic under the randomisation hypothesis

-   Var.Ii: the variance of local Moran statistic under the randomisation hypothesis

-   Z.Ii:the standard deviate of local Moran statistic

-   Pr(): the p-value of local Moran statistic

## Mapping Local Moran's I values & P-values

Append the local Moran's I dataframe (i.e. localMI) onto total_trips_by_grid spatialPolygonDataFrame.

```{r}
ttbg_sf.localMI_wdmp <- cbind(total_trips_by_grid_wdmp_sf,localMI_wdmp) %>%
  rename(Pr.Ii = Pr.z....E.Ii..)

ttbg_sf.localMI_wdap <- cbind(total_trips_by_grid_wdap_sf,localMI_wdap) %>%
  rename(Pr.Ii = Pr.z....E.Ii..)

ttbg_sf.localMI_hmp <- cbind(total_trips_by_grid_hmp_sf,localMI_hmp) %>%
  rename(Pr.Ii = Pr.z....E.Ii..)

ttbg_sf.localMI_hep <- cbind(total_trips_by_grid_hep_sf,localMI_hep) %>%
  rename(Pr.Ii = Pr.z....E.Ii..)
```

## Creating a LISA Cluster Map

Shows the significant locations colour coded by type of spatial autocorrelation.

### **Preparing LISA Map Classes**

Create an empty numeric vector named quadrant with a length equal to the number of rows in the object localMI

```{r}
quadrant_wdmp <- vector(mode="numeric",length=nrow(localMI_wdmp))

quadrant_wdap <- vector(mode="numeric",length=nrow(localMI_wdap))

quadrant_hmp <- vector(mode="numeric",length=nrow(localMI_hmp))

quadrant_hep <- vector(mode="numeric",length=nrow(localMI_hep))
```

Setting Significance Level

```{r}
signif <- 0.05
```

#### **Weekday Morning Peak 6am-9am**

```{r}

# Derives the spatially lagged variable of interest (i.e. Total Trips) and centers the spatially lagged variable around its mean

total_trips_by_grid_wdmp_sf$total_trips_lag <- lag.listw(rswm_wdmp, total_trips_by_grid_wdmp_sf$total_trips)
DV_wdmp <- total_trips_by_grid_wdmp_sf$total_trips_lag - mean(total_trips_by_grid_wdmp_sf$total_trips_lag)

#centering the local Moran’s around the mean

LM_I_wdmp <- localMI_wdmp[,1]

#The four command lines define the low-low (1), low-high (2), high-low (3) and high-high (4) categories.

quadrant_wdmp[DV_wdmp <0 & LM_I_wdmp>0] <- 1
quadrant_wdmp[DV_wdmp >0 & LM_I_wdmp<0] <- 2
quadrant_wdmp[DV_wdmp <0 & LM_I_wdmp<0] <- 3  
quadrant_wdmp[DV_wdmp >0 & LM_I_wdmp>0] <- 4    

#non-significant Moran placed in the category 0

quadrant_wdmp[localMI_wdmp[,5]>signif] <- 0
```


### **Plotting LISA map**

Plot both the local Moran's I values map and its corresponding p-values map next to each other for effective interpretation

#### **Weekday Morning Peak 6am-9am**

```{r}

tmap_mode("plot")

total_trips_wdmp <- qtm(ttbg_sf.localMI_wdmp, "total_trips")

ttbg_sf.localMI_wdmp$quadrant <- quadrant_wdmp

colors <- c("#ffffff", "#2c7bb6", "#abd9e9", "#fdae61", "#d7191c")
clusters <- c("insignificant", "low-low", "low-high", "high-low", "high-high")

LISAmap_wdmp <- tm_shape(ttbg_sf.localMI_wdmp) +
  tm_fill(col = "quadrant", 
          style = "cat", 
          palette = colors[c(sort(unique(quadrant_wdmp)))+1], 
          labels = clusters[c(sort(unique(quadrant_wdmp)))+1],
          alpha = 0.6) +
  tm_view(set.zoom.limits = c(11,17)) +
  tm_borders(alpha=0.5)

tmap_arrange(total_trips_wdmp, LISAmap_wdmp, 
             asp=1, ncol=2)
```

::: {.callout-tip title="Statistical Conclusion" style="color: gray"}
Weekday Morning Peak 6am-9am

Of the hexagons that are significant, there are only regions that are:

-   Low-High (low values of total travels but have neighbors that also have high values of total travels) - contribute significantly to a positive global spatial autocorrelation outcome

-   High-High (high values of total travels and have neighbors that also have high values of total travels) - contribute significantly to a negative global spatial autocorrelation outcome

High-high hexagons seem to be concentrated at residential areas e.g. Boon Lay, Bukit Batok, while Low-High hexagons seem to be around industrial/office areas e.g. International Road, CBD
:::
