---
title: "Take Home Exercise 1"
date: "27 Novemeber 2023"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  warning: false
editor: visual
---

## Getting Started

```{r}
pacman::p_load(tmap,sf,tidyverse,knitr,h3jsr,dplyr, mapview)
```

## Preparing the Flow Data

### Importing the OD data

Firstly, we will import the **Passenger Volume by Origin Destination Bus Stops** data set downloaded from LTA DataMall by using read_csv() of **readr** package.

```{r}
odbus <- read_csv("data/aspatial/origin_destination_bus_202308.csv")
```

Check odbus tibble data frame that values in OROGIN_PT_CODE and DESTINATION_PT_CODE are in numeric data type.

```{r}
glimpse(odbus)
```

Origin & Destination Bus Stop Code

```{r}
odbus$ORIGIN_PT_CODE <-
as.factor(odbus$ORIGIN_PT_CODE)
odbus$DESTINATION_PT_CODE <-
as.factor(odbus$DESTINATION_PT_CODE)
```

### Extracting the Study Data

Filter out data that belong to trips that occur on:

-   "Weekday" and "6-9am" (wdmp)

    ```{r}
    wdmp <-  odbus %>%
      filter(DAY_TYPE == "WEEKDAY") %>%
      filter(TIME_PER_HOUR >= 6 &
               TIME_PER_HOUR <= 9) %>%
      group_by(ORIGIN_PT_CODE) %>%
      summarise(TRIPS = sum(TOTAL_TRIPS))
    ```

-   "Weekday" and "5-8pm" (wdap)

    ```{r}
    wdap <-  odbus %>%
      filter(DAY_TYPE == "WEEKDAY") %>%
      filter(TIME_PER_HOUR >= 17 &
               TIME_PER_HOUR <= 20) %>%
      group_by(ORIGIN_PT_CODE) %>%
      summarise(TRIPS = sum(TOTAL_TRIPS))
    ```

-   "Weekends/Holiday" and "11am-2pm" (hmp)

    ```{r}
    hmp <- odbus %>%
      filter(DAY_TYPE == "WEEKENDS/HOLIDAY") %>%
      filter(TIME_PER_HOUR >= 11 &
               TIME_PER_HOUR <= 14) %>%
      group_by(ORIGIN_PT_CODE) %>%
      summarise(TRIPS = sum(TOTAL_TRIPS))
    ```

-   "Weekends/Holiday" and "4pm-7pm" (hep)

    ```{r}
    hep <- odbus %>%
      filter(DAY_TYPE == "WEEKENDS/HOLIDAY") %>%
      filter(TIME_PER_HOUR >= 16 &
               TIME_PER_HOUR <= 19) %>%
      group_by(ORIGIN_PT_CODE) %>%
      summarise(TRIPS = sum(TOTAL_TRIPS))
    ```

Check resulting data tables

```{r}
kable(head(wdmp)) 
kable(head(wdap)) 
kable(head(hmp)) 
kable(head(hep))
```

Output saved in rds format for future use

```{r}
write_rds(wdmp, "data/rds/wdmp.rds") 
write_rds(wdap, "data/rds/wdap.rds") 
write_rds(hmp, "data/rds/hmp.rds") 
write_rds(hep, "data/rds/hep.rds")
```

Import the rds file into R environment

```{r}
wdmp <- read_rds("data/rds/wdmp.rds") 
wdap <- read_rds("data/rds/wdap.rds") 
hmp <- read_rds("data/rds/hmp.rds") 
hep <- read_rds("data/rds/hmp.rds")
```

## Working with Geospatial Data

Two geospatial data (shapefile) will be used for this exercise:

-   BusStop: Provides location of bus stop as at Q4 2022

-   MPSZ-2019: This data provides the sub-zone boundary of URA Master Plan 2019

### Importing geospatial data

```{r}
busstop <- st_read(dsn = "Data/geospatial",
                   layer = "BusStop") %>%
  st_transform(crs = 3414)

mpsz <- st_read(dsn = "data/geospatial",
                layer = "MPSZ-2019") %>%
  st_transform(crs = 3414)
```

Check structure of `busstop` and `MPSZ`  sf tibble data frame

```{r}
glimpse(busstop)
glimpse(mpsz)
```

## Geospatial Data Wrangling

## Combining Busstop & mpsz

```{r}
busstop_mpsz <- st_intersection(busstop, mpsz) %>%
  select(BUS_STOP_N, SUBZONE_C) %>%
  st_drop_geometry()
```

Save output into rds format

```{r}
write_rds(busstop_mpsz, "data/rds/busstop_mpsz.csv")
```

Each data frame will have the planning subzone code appended to it, and only unique records will be retained.

-   Weekday Morning Peak

    ```{r}
    wdmp_c <- left_join(wdmp , busstop_mpsz,
                by = c("ORIGIN_PT_CODE" = "BUS_STOP_N")) %>%
      
      rename(ORIGIN_BS = ORIGIN_PT_CODE,
             ORIGIN_SZ = SUBZONE_C) 

    duplicate <- wdmp_c %>%
      group_by_all() %>%
      filter(n()>1) %>%
      ungroup() 

    wdmp_o <- unique(wdmp_c)
    ```

-   Weekday Afternoon Peak

    ```{r}
    wdap_c <- left_join(wdap , busstop_mpsz,
                by = c("ORIGIN_PT_CODE" = "BUS_STOP_N")) %>%
      
      rename(ORIGIN_BS = ORIGIN_PT_CODE,
             ORIGIN_SZ = SUBZONE_C) 

    duplicate <- wdap_c %>%
      group_by_all() %>%
      filter(n()>1) %>%
      ungroup()  

    wdap_o <- unique(wdap_c)
    ```

-   Weekends/Holiday Morning Peak

    ```{r}
    hmp_c <- left_join(hmp , busstop_mpsz,
                by = c("ORIGIN_PT_CODE" = "BUS_STOP_N")) %>%
      
      rename(ORIGIN_BS = ORIGIN_PT_CODE,
             ORIGIN_SZ = SUBZONE_C) 

    duplicate <- hmp_c %>%
      group_by_all() %>%
      filter(n()>1) %>%
      ungroup() 

    hmp_o <- unique(hmp_c)
    ```

-   Weekends/Holiday Evening Peak

    ```{r}
    hep_c <- left_join(hep , busstop_mpsz,
                by = c("ORIGIN_PT_CODE" = "BUS_STOP_N")) %>%
      
      rename(ORIGIN_BS = ORIGIN_PT_CODE,
             ORIGIN_SZ = SUBZONE_C) 

    duplicate <- hep_c %>%
      group_by_all() %>%
      filter(n()>1) %>%
      ungroup() 

    hep_o <- unique(hep_c)
    ```

Update data with the planning subzone codes

```{r}
mpsz_wdmp_o <- left_join(mpsz, 
                       wdmp_o,
                       by = c("SUBZONE_C" = "ORIGIN_SZ"))

mpsz_wdap_o <- left_join(mpsz, 
                       wdap_o,
                       by = c("SUBZONE_C" = "ORIGIN_SZ"))

mpsz_hmp_o <- left_join(mpsz, 
                      hmp_o,
                      by = c("SUBZONE_C" = "ORIGIN_SZ"))

mpsz_hep_o <- left_join(mpsz, 
                      hep_o,
                      by = c("SUBZONE_C" = "ORIGIN_SZ"))


```

## **Choropleth Visualisation**

Choropleth map showing the distribution of passenger trips at planning sub-zone level.

```{r}
tm_shape(mpsz_hep_o)+
  tm_fill("TRIPS", 
          style = "quantile", 
          palette = "Greens",
          title = "Passenger trips") +
  tm_layout(main.title = "Passenger trips generated at planning sub-zone level",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from URA\n and Passenger trips data from LTA", 
             position = c("left", "bottom"))
```
