---
title: "Hands-on Exercise 1: Geospatial Data Wrangling with R"
editor: visual
---

## Overview

In this hands-on exercise, I learned how to import and wrangle geospatial data using appropriate R packages.

## Getting Started

The code chunk below installs and load sf and tidyverse packages into R environment

```{r}
pacman::p_load(sf, tidyverse)
```

## Importing Geospatial Data

### Importing polygon feature data

```{r}
mpsz <- st_read(dsn = "Data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

### **Importing polyline feature data in shapefile form**

The code chunk below uses *st_read()* function of **sf** package to import `CyclingPath` shapefile into R as line feature data frame.

```{r}
cyclingpath = st_read(dsn = "Data/geospatial", 
                         layer = "CyclingPathGazette")
```

### **Importing GIS data in kml format**

The `pre-schools-location-kml` is in kml format. The code chunk below will be used to import the kml into R.

```{r}
preschool = st_read("data/geospatial/preschoolslocation.kml")
```

## **Checking the Content of A Simple Feature Data Frame**

In this sub-section, I learned different ways to retrieve information related to the content of a simple feature data frame.
